NOAA Severe Weather Events - Economic and Human Effects
---
##Synopsis
Data from the US National Oceanic and Atmospheric Administration's 
(NOAA) database was used to analyze the economic and human impacts of different 
weather events. While tornados overwhelming cause the most injuries and deaths, 
floods and coastal storm surges caused the largest amount of property and crop 
damage. This conclusion is skewed by the huge economic impact of Hurricane 
Katrina in 2005. 

##Data Processing
Data was taken from the US National Oceanic and Atmospheric Administration's 
(NOAA) database.  
```{r, echo=TRUE, cache=TRUE}
setwd(paste('/home/tom/scripts/gnuR/coursera/datasciencecoursera/',
      'reproducible_research/peer_assgn_2', sep=""))
if(!file.exists('data')){
    dir.create('data')
}
file_url = paste('https://d396qusza40orc.cloudfront.net',
                 '/repdata%2Fdata%2FStormData.csv.bz2', sep="")
download.file(file_url, destfile = './data/NOAA_data.bz2', method="curl")
#Note: a .bz2 unzip utility could not be found. The file must be manually
#unzipped into the `working_directory/data` folder
NOAA <- read.csv('./data/NOAA_data', header=TRUE)
```

The event types were adjusted and homogenized using text matching. 
Those that didn't fit into a 
category were classified as "Other," and were later evaluated to check the size
and importance of that category. It turned out that, of 18 events classified 
and one "Other" category, the 10th largest event by economic impact is 
approximately the same size as the last 9 categories aggregated. This implies 
that a sufficient amount of categories were described in order to catch the 
most important types of weather event.  
  
The property and crop damage exponents were multiplied with the given values 
to get the real monetary values. A few cases (around 400) had unknown exponent 
values, hence these were discarded.

```{r, echo=TRUE, cache=TRUE}
evs <- tolower(NOAA$EVTYPE)
events <- rep("Other", length(evs))
events[c(grep("flood", evs), grep("flooding", evs), grep("rising water", evs), 
      grep("floood", evs))] <- "Flood"
events[c(grep("tornado", evs), grep("torndao", evs))] <- "Tornado"
events[c(grep("hail", evs), grep("freezing", evs), grep("ice", evs),
      grep("glaze", evs), grep("icy", evs))] <- 
    "Hail/Freezing Rain/Ice"
events[c(grep("winter", evs), grep("snow", evs), grep("sleet", evs),
      grep("blizzard", evs), grep("wintry", evs), grep("mix", evs))] <- 
    "Winter/Snow/Blizzard"
events[c(grep("tide", evs), grep("coastal", evs), grep("surf", evs), 
      grep("sea", evs), grep("wave", evs), grep("surge", evs), 
      grep("swell", evs))] <- "Tide/Wave Event"
events[c(grep("summary", evs), grep("remnant", evs))] <- NA
events[c(grep("hurricane", evs), grep("tsunami", evs), grep("tropical", evs), 
      grep("typhoon", evs))] <- "Hurricane/Tsunami/Tropical Storm"
events[c(grep("wind", evs), grep("wnd", evs))] <- "Wind"
events[c(grep("dry", evs), grep("drought", evs))] <- "Dry/Drought"
events[c(grep("cold", evs), grep("frost", evs), grep("cool", evs),
      grep("freeze", evs), grep("low temperature", evs), 
      grep("record low", evs))] <- "Cold"
events[grep("volcan", evs)] <- "Volcanic Ash"
events[grep("fire", evs)] <- "Wildfire"
events[c(grep("thunderstorm", evs), grep("lightning", evs), 
      grep("lighting", evs), grep("ligntning", evs))] <- "Thunderstorms"
events[c(grep("mud", evs), grep("rock", evs), grep("landslide", evs), 
      grep("landslump", evs))] <- "Mudslide/Rockslide/Landslide"
events[c(grep("avalance", evs), grep("avalanche", evs))] <- "Avalanche"
events[c(grep("warm", evs), grep("heat", evs), grep("hot", evs), 
      grep("record high", evs), grep("hyperthermia", evs), 
      grep("high temperature", evs))] <- "Heat"
events[c(grep("rain", evs), grep("wet", evs), grep("precipitation", evs), 
      grep("precipatation", evs), grep("showers", evs))] <- "Rain"
events[c(grep("fog", evs), grep("vog", evs))] <- "Fog"
events[c(grep("spout", evs))] <- "Waterspouts"

prop.dmg <- NOAA$PROPDMG
prop.dmg[prop.dmg==""] <- 0
prop.factor <- rep(NA, length(prop.dmg))
prop.factor[NOAA$PROPDMGEXP==""] <- 1
prop.factor[((NOAA$PROPDMGEXP=="k") | (NOAA$PROPDMGEXP=="K"))] <- 1e3
prop.factor[((NOAA$PROPDMGEXP=="m") | (NOAA$PROPDMGEXP=="M"))] <- 1e6
prop.factor[((NOAA$PROPDMGEXP=="b") | (NOAA$PROPDMGEXP=="B"))] <- 1e9
#There is a serious error in this reference to the Napa Valley floods in
#January 2006
prop.factor[(NOAA$REFNUM==605943)] <- 1e6
prop.dmg <- prop.dmg*prop.factor/1e6

crop.dmg <- NOAA$CROPDMG
crop.dmg[prop.dmg==""] <- 0
crop.factor <- rep(NA, length(crop.dmg))
crop.factor[NOAA$CROPDMGEXP==""] <- 1
crop.factor[((NOAA$CROPDMGEXP=="k") | (NOAA$CROPDMGEXP=="K"))] <- 1e3
crop.factor[((NOAA$CROPDMGEXP=="m") | (NOAA$CROPDMGEXP=="M"))] <- 1e6
crop.factor[((NOAA$CROPDMGEXP=="b") | (NOAA$CROPDMGEXP=="B"))] <- 1e9
crop.dmg <- crop.dmg*crop.factor/1e6

date <- strptime(NOAA$BGN_DATE, "%m/%d/%Y %H:%M:%S")

NOAA.clean <- data.frame("Date"=date, "State"=NOAA$STATE, "Event"=events, 
                         "Fatalities"=NOAA$FATALITIES, 
                         "Injuries"=NOAA$INJURIES, "Property.Damage"=prop.dmg, 
                         "Crop.Damage"=crop.dmg, 
                         "Year"=as.numeric(format(date, "%Y")))
NOAA.clean <- NOAA.clean[complete.cases(NOAA.clean), ]
remove(NOAA)
```

##Results: Human Impact
Total casualties (deaths plus injuries) were analyzed across 
the different weather event types.
```{r, cache=TRUE}
total.casualties <- aggregate((NOAA.clean$Fatalities + 
                                 NOAA.clean$Injuries), 
                           by=list(NOAA.clean$Event, NOAA.clean$Year), 
                           FUN=sum)
names(total.casualties) <- c("Event", "Year", "Total.Casualties")

library(ggplot2)
qplot(Year, Total.Casualties, data=total.casualties, color=Event, geom="line", 
      ylab="Total Casualties", main="Total Casualties Related to Weather")
```

If the total casualties are summed across the recorded years, the effects of 
other weather events are easier to perceive.
```{r, cache=TRUE}
casualties.totals <- aggregate((NOAA.clean$Fatalities + NOAA.clean$Injuries),
                         by=list(NOAA.clean$Event), FUN=sum)
names(casualties.totals) <- c("Event", "Total.Casualties")
casualties.totals<- casualties.totals[order(casualties.totals$Total.Casualties, 
                                             decreasing=TRUE), ]
print(casualties.totals)
```

##Results: Economic Impact
Crop damage and property damage can be summed together for a total economic 
impact estimate.  
    
It's important to note, for instance, that a large amount of damage from 
Hurrican Katrina in 2005, namely $31.3 billion, is classified as 
"Storm Surge", which I reclassified as "Tide/Wave Event"
```{r, cache=TRUE}
total.dmg.year <- aggregate((NOAA.clean$Property.Damage + 
                                 NOAA.clean$Crop.Damage), 
                           by=list(NOAA.clean$Event, NOAA.clean$Year), 
                           FUN=sum)
names(total.dmg.year) <- c("Event", "Year", "Total.Damage")

qplot(Year, Total.Damage, data=total.dmg.year, color=Event, geom="line", 
      ylab="Total Damage (in Million USD)", xlim=c(1990, 2015))
```

If the total damage is summed across the recorded years, the effects of other 
weather events are easier to perceive.
```{r}
hist.totals <- aggregate((NOAA.clean$Property.Damage + NOAA.clean$Crop.Damage),
                         by=list(NOAA.clean$Event), FUN=sum)
names(hist.totals) <- c("Event", "Total.Damage")
hist.totals <- hist.totals[order(hist.totals$Total.Damage, decreasing=TRUE), ]
print(hist.totals)
```

The top 10 largest contributors are shown below.
```{r}
other.events <- sum(hist.totals$Total.Damage[11:dim(hist.totals)[1]])
top.damage <- data.frame(c(hist.totals$Total.Damage[1:10], other.events),
                           row.names=c(as.character(hist.totals$Event[1:10]), 
                                       "All Other"))
names(top.damage) <- "Damage"
pie(top.damage$Damage, labels=row.names(top.damage), 
    main="Total Damage (in Million USD)", col=rainbow(11))
```

##Conclusions
Clearly, tornados cause the largest human impact. The results in terms of 
economic impact are less clear, as one single event, Hurricane Katrina, caused 
a massive amount of damage. However, it is not trivial to remove this event 
from the existing dataset, as the entries are mixed into different categories 
such as "Storm Surge," "Hurricane," and possibly others.